Gravitational field calculation in spherical coordinates using variable densities in depth

### 2019-02-12 manuscript/manuscript.tex
Add a version with differences highlighted
Used git latexdiff to create it and include here in case we don't have
it installed.
Fix wrong or missing punctuation on equations
Fix typos highlighted by reviewer

\begin{equation}
    g_{\alpha}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') \frac{\Delta_\alpha}{\ell^3}
    \kappa \, dr' d\phi' d\lambda',
\label{eq:tesseroid-grav}
\end{equation}
\\
\noindent in which $\alpha \in \{x, y, z\}$,
$G = 6.674\times10^{-11}\, \text{m$^3$kg$^{-1}$s$^{-1}$}$ is the gravitational constant
and

@@

\iftwocol{
\begin{equation}
    \begin{split}
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \rho(r') f(r', \phi', \lambda')
        dr' d\phi' d\lambda' \approx& \\
        A
        \sum\limits_{i=1}^{N^r}
        \sum\limits_{j=1}^{N^\phi}
        \sum\limits_{k=1}^{N^\lambda}
        W_i^r W_j^\phi W_k^\lambda
        &\rho(r_i) f(r_i, \phi_j, \lambda_k),
    \end{split}
\label{eq:glq-var-dens}
\end{equation}
}{
\begin{equation}
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') f(r', \phi', \lambda')
    dr' d\phi' d\lambda' \approx
    A
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    W_i^r W_j^\phi W_k^\lambda \rho(r_i) f(r_i, \phi_j, \lambda_k),
\label{eq:glq-var-dens}
\end{equation}
}
\\
\noindent where

\begin{equation}
    A = \frac{(\lambda_2 - \lambda_1)(\phi_2 - \phi_1)(r_2 - r_1)}{8},
\end{equation}
\\
\noindent $f(r', \phi', \lambda')$ is an integral kernel for a homogeneous tesseroid
\citep{Grombein2013},
$(r_i, \phi_j, \lambda_k)$ are the coordinates of the quadrature nodes,
$N^r$, $N^\phi$, $M^\lambda$ are the quadrature orders and $W_i^r$, $W_j^\phi$,
$W_k^\lambda$ are the quadrature weights in the radial, latitudinal, and longitudinal
directions, respectively.
\citet[p.~391]{Hildebrand1987} provides the formulas for calculating the GLQ weights and
pre-computed values of the Legendre polynomial roots for low orders.
See \citet{Uieda2016} for a more detailed description using a similar notation to the
one used here.
It is worth noting that a GLQ is equivalent to approximating the tesseroid by $N^r
\times N^\phi \times N^\lambda$ point masses located on the quadrature nodes
\citep{Ku1977, Asgharzadeh2007}.

@@

\begin{equation}
    \cos\psi_t =
        \sin\phi\sin\phi_t + \cos\phi\cos\phi_t\cos(\lambda - \lambda_t).
\end{equation}

\begin{equation}
    r_t = \frac{r_2 + r_1}{2}, \quad
    \phi_t = \frac{\phi_2 + \phi_1}{2}, \quad
    \lambda_t = \frac{\lambda_2 + \lambda_1}{2}.
\end{equation}
\\
\noindent
and the dimensions of the tesseroid are defined as

\begin{equation}
    L_\lambda = r_2 \arccos(\sin^2\phi_t +
        \cos^2\phi_t\cos(\lambda_2 - \lambda_1)),
    \label{eq:sizelon}
\end{equation}
\\
\noindent and

@@

\begin{equation}
    \rho_l(r') =
    \frac{ \rho_n(r_2) - \rho_n(r_1) }{ r_2 - r_1 } (r' - r_1) + \rho_n(r_1),
    \label{eq:density-reference-line}
\end{equation}

@@

On the first iteration, the ratio $L_r/L_r^\text{orig} = 1$ because the tesseroid being
divided is the original one.
For further iterations, the ratio will be progressively smaller than one as the
tesseroids get smaller.
This is intended to limit the number of divisions to the ones that will
significantly reduce the numerical error:
dividing a large tesseroid with a small $\text{max}\{ \Delta \rho(r') \}$ would
improve the integration accuracy more than dividing a small tesseroid with a
higher $\text{max}\{ \Delta \rho(r') \}$.

@@

We empirically determined optimal values for $D$ and $\delta$ by comparing the numerical
results with analytical solutions.
Our analysis included a range of tesseroid models and computation grids as well as
common density functions.
These values minimize the computational load while maintaining the numerical error below
0.1\% of an analytical solution.
The density functions used to establish the optimal values were a linear, an
exponential, and a sinusoidal function.
The linear function represents the smoothest variation of density and does not require
density-based discretization at all.
Through this analysis, we have obtained optimal values for the distance-size ratio $D$
of 1 and 2.5 for the potential an its gradient, respectively.
We analysed the error for exponential functions ranging from smooth to sharp and
sinusoidal function of different wavelengths to test the accuracy of the density-based
discretization.
Values of $\delta$ equal to 0.1 are sufficient to guarantee a 0.1\% accuracy for most
realist cases tested.
These results could be extrapolated to other continuous density functions that are
sufficiently smooth, as is the case for most common geophysical applications.
Smaller values of $\delta$ can used for more highly variable density functions to
increase the accuracy of the integration accordingly.

@@

\begin{equation}
    g_z(r) = \frac{V_\text{sh}(r)}{r},
\label{eq:shell-gz}
\end{equation}

@@

\begin{equation}
    \begin{split}
    V_\text{sine}(r) = \frac{4\pi G}{r} \frac{A}{k^3} \Big[
    & (2 - k^2 R_2^2) \cos(k(R_2 - R)) + 2 k R_2 \sin(k(R_2 - R)) - \\
    & (2 - k^2 R_1^2) \cos(k(R_1 - R)) - 2 k R_1 \sin(k(R_1 - R))
        \Big]
    \end{split}
\end{equation}

>>>>

\begin{equation}
    g_{\alpha}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') \frac{\Delta_\alpha}{\ell^3}
    \kappa \, dr' d\phi' d\lambda',
\label{eq:tesseroid-grav}
\end{equation}

\noindent in which $\alpha \in \{x, y, z\}$,
$G = 6.674\times10^{-11}\, \text{m$^3$kg$^{-1}$s$^{-1}$}$ is the gravitational constant
and

@@

\iftwocol{
\begin{equation}
    \begin{split}
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \rho(r') f(r', \phi', \lambda')
        dr' d\phi' d\lambda' \approx& \\
        A
        \sum\limits_{i=1}^{N^r}
        \sum\limits_{j=1}^{N^\phi}
        \sum\limits_{k=1}^{N^\lambda}
        W_i^r W_j^\phi W_k^\lambda
        &\rho(r_i) f(r_i, \phi_j, \lambda_k),
    \end{split}
\label{eq:glq-var-dens}
\end{equation}
}{
\begin{equation}
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') f(r', \phi', \lambda')
    dr' d\phi' d\lambda' \approx
    A
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    W_i^r W_j^\phi W_k^\lambda \rho(r_i) f(r_i, \phi_j, \lambda_k),
\label{eq:glq-var-dens}
\end{equation}
}

\begin{equation}
    A = \frac{(\lambda_2 - \lambda_1)(\phi_2 - \phi_1)(r_2 - r_1)}{8},
\end{equation}

\noindent $f(r', \phi', \lambda')$ is an integral kernel for a homogeneous tesseroid
\citep{Grombein2013},
$(r_i, \phi_j, \lambda_k)$ are the coordinates of the quadrature nodes,
$N^r$, $N^\phi$, $M^\lambda$ are the quadrature orders and $W_i^r$, $W_j^\phi$,
$W_k^\lambda$ are the quadrature weights in the radial, latitudinal, and longitudinal
directions, respectively.
\citet[p.~391]{Hildebrand1987} provides the formulas for calculating the GLQ weights and
pre-computed values of the Legendre polynomial roots for low orders.
See \citet{Uieda2016} for a more detailed description using a similar notation to the
one used here.
It is worth noting that a GLQ is equivalent to approximating the tesseroid by $N^r
\times N^\phi \times N^\lambda$ point masses located on the quadrature nodes
\citep{Ku1977, Asgharzadeh2007}.

@@

\begin{equation}
    \cos\psi_t =
        \sin\phi\sin\phi_t + \cos\phi\cos\phi_t\cos(\lambda - \lambda_t),
\end{equation}

\begin{equation}
    r_t = \frac{r_2 + r_1}{2}, \quad
    \phi_t = \frac{\phi_2 + \phi_1}{2}, \quad
    \lambda_t = \frac{\lambda_2 + \lambda_1}{2},
\end{equation}

\noindent
and the dimensions of the tesseroid are defined as

\begin{equation}
    L_\lambda = r_2 \arccos(\sin^2\phi_t +
        \cos^2\phi_t\cos(\lambda_2 - \lambda_1)),
    \label{eq:sizelon}
\end{equation}

@@

\begin{equation}
    \rho_l(r') =
    \frac{ \rho_n(r_2) - \rho_n(r_1) }{ r_2 - r_1 } (r' - r_1) + \rho_n(r_1).
    \label{eq:density-reference-line}
\end{equation}

@@

On the first iteration, the ratio $L_r/L_r^\text{orig}$ equals one because the tesseroid
being divided is the original one.
For further iterations, the ratio will be progressively smaller than one as the
tesseroids get smaller.
This is intended to limit the number of divisions to the ones that will
significantly reduce the numerical error:
dividing a large tesseroid with a small $\text{max}\{ \Delta \rho(r') \}$ would
improve the integration accuracy more than dividing a small tesseroid with a
higher $\text{max}\{ \Delta \rho(r') \}$.

@@

We empirically determined optimal values for $D$ and $\delta$ by comparing the numerical
results with analytical solutions.
Our analysis included a range of tesseroid models and computation grids as well as
common density functions.
These values minimize the computational load while maintaining the numerical error below
0.1\% of an analytical solution.
The density functions used to establish the optimal values were a linear, an
exponential, and a sinusoidal function.
The linear function represents the smoothest variation of density and does not require
density-based discretization at all.
Through this analysis, we have obtained optimal values for the distance-size ratio $D$
of 1 and 2.5 for the potential and its gradient, respectively.
We analysed the error for exponential functions ranging from smooth to sharp and
sinusoidal function of different wavelengths to test the accuracy of the density-based
discretization.
Values of $\delta$ equal to 0.1 are sufficient to guarantee a 0.1\% accuracy for most
realist cases tested.
These results could be extrapolated to other continuous density functions that are
sufficiently smooth, as is the case for most common geophysical applications.
Smaller values of $\delta$ can be used for more highly variable density functions to
increase the accuracy of the integration accordingly.

@@

\begin{equation}
    g_z(r) = \frac{V_\text{sh}(r)}{r}.
\label{eq:shell-gz}
\end{equation}

@@

\begin{equation}
    \begin{split}
    V_\text{sine}(r) = \frac{4\pi G}{r} \frac{A}{k^3} \Big[
    & (2 - k^2 R_2^2) \cos(k(R_2 - R)) + 2 k R_2 \sin(k(R_2 - R)) - \\
    & (2 - k^2 R_1^2) \cos(k(R_1 - R)) - 2 k R_1 \sin(k(R_1 - R))
        \Big].
    \end{split}
\end{equation}
